---
title: "Buybacks & Investment"
author: "Hamdi Ahmed"
date: '2018-08-12'
slug: buybacks-investment
categories: []
tags: []
---

Since the Trump Tax cuts were approved, there's been a surge in buybacks and a fall in tax revenues from corporations.




``` {r include = FALSE, message = FALSE, echo = FALSE}
library(dplyr)
library(plotly)
library(tidyverse)
library(readxl)
library(lubridate)

plot <- read_xls("buyback.xls")

plot$observation_date <- as.Date(plot$observation_date)

accumulate_by <- function(dat, var) {
  var <- lazyeval::f_eval(var, dat)
  lvls <- plotly:::getLevels(var)
  dats <- lapply(seq_along(lvls), function(x) {
    cbind(dat[var %in% lvls[seq(1, x)], ], frame = lvls[[x]])
  })
  dplyr::bind_rows(dats)
}

f <- plot %>%
  accumulate_by(~observation_date)


d <- ggplot(f, aes(observation_date, frame = frame)) +
  geom_line(aes(y = net_buyback_gdp, colour = "net buybacks"), size = 1) + 
  geom_line(aes(y = corporate_tax_gdp, colour = "corporate tax income"), size = 1) +
  geom_vline(xintercept = as.numeric(plot$observation_date[20]), linetype = "dashed", ) +
  annotate("text", label = "Tax Cut Approved", x = plot$observation_date[15], y = 0, color = "black") +
  labs(y = "Percent of GDP", x = "Date") +
  theme(legend.title = element_blank())
  
```

``` {r include = FALSE, message = FALSE, echo = FALSE}
p <- ggplotly(d, width = 800) %>%
  layout(legend=list(y=0.8, yanchor="top")) %>% 
  add_annotations( text="Colour", xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.8, yanchor="bottom",    # Same y as legend below
                  legendtitle=TRUE, showarrow=FALSE ) %>%
   animation_opts(
    frame = 100,
    easing = "linear",
    transition = 0, 
    redraw = FALSE
  ) %>% 
  animation_slider(
    currentvalue = list(
      prefix = "Date"
    )
  )


```

```{r echo = FALSE}
p
```

It's fine though since corporate tax [cuts](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1091238) are supposed to [lead](https://twitter.com/realdonaldtrump/status/942123433873281024?lang=en) to [higher](https://www.bloomberg.com/view/articles/2017-11-17/yes-a-corporate-tax-cut-would-increase-investment) business investment but there’s claims that companies are foregoing investment and using the extra cash to buyback their [shares](http://theweek.com/articles/787966/ban-share-buybacks?utm_source=dlvr.it&utm_medium=twitter). Indeed, buybacks haven’t just been rising due to the recent tax cut, they have been on the rise since they became legal in 1982 and corporate capital investment, although cyclical, has been trending downwards since then.


``` {r include = FALSE, message = FALSE, echo = FALSE}

library(zoo)
library(tidyr)


invest <- read_xlsx("invest.xlsx")

invest$year <- as.Date(as.yearqtr(invest$year))



m <- ggplot(invest, aes(x = year)) +
  geom_line(aes(y = net_buybacks, colour = "Net buybacks")) + 
  geom_line(aes(y = investment, colour = "U.S Investment")) +
  labs(y = "Percent of GDP", x = "Year") 

```

```{r echo = FALSE, fig.align='center'}

m

```
Corporations having more cash at hand and not increasing their investments doesn’t bode well for the economy. This is understandable though since capital expenditure needs careful planning before it’s undertaken and so we may see this rise in the future. Capital expenditure plans are also affected by uncertainty and things like a possible trade war may be delaying [plans](https://www.telegraph.co.uk/business/2018/06/20/us-businesses-putting-investment-trade-war-worries-fed-chief/). All of this has led me to wonder, do share buybacks deter investment?

First, some definitions. Share buybacks or share repurchases is the act of returning cash to shareholders by buying stock held by them. Also, with less shares available, the remaining investors get a bigger piece of corporate profits. However, corporations may issue new shares in the same time period as they buy back shares, so I use net share buybacks in my graphs to take this into account. Net share buybacks also show us the amount that is left over for investment. I use investment defined as gross fixed capital formation undertaken by the non-financial corporate sector. Gross fixed capital formation measures the net increase in fixed capital investment which is a component of the [expenditure method in calculating GDP](http://www.econport.org/content/handbook/NatIncAccount/CalculatingGDP/Expenditures.html). Both data sets are from the [Federal Reserve](https://www.federalreserve.gov/econres/notes/ifdp-notes/corporate-buybacks-and-capital-investment-an-international-perspective-20170411.htm).

If a company is buying back more shares then there’s less resources available to devote to investment right?  John Cochrane says otherwise as companies could be giving back cash to shareholders because of a lack of profitable projects to [invest in](https://johnhcochrane.blogspot.com/2017/12/the-buyback-fallacy.html).

>If company 1 doesn't have any good investment ideas, even after the tax cut, and company 2 does have some good investment ideas, made better after the tax cut, the economy needs to get money from company 1 to company 2. Company 1 could buy company 2; company 1 could invest in company 2 by buying its stock or buying its debt (all that "cash" you hear about); company 1 could return money to shareholders, and the shareholders could invest in company 2. They're all the same, to economics. Of all the ways to do this, actually, the last might well be the most efficient. Shareholders might have better ideas about good investments than managers of a company that doesn't have any good investment ideas.

It becomes an issue though if most companies are handing back cash to [shareholders](https://www.bloomberg.com/view/articles/2018-07-18/trump-s-tax-cut-hasn-t-done-anything-for-workers). Is this a sign of a lack of good investment projects available in the economy or would companies be investing more if they weren’t distracted by the ability to buyback shares?
