---
title: "Buybacks & Investment"
author: "Hamdi Ahmed"
date: '2018-08-12'
slug: buybacks-investment
categories: []
tags: []
---

Since the Trump Tax cuts were approved, there's been a surge in buybacks and a fall in tax revenues from corporations.




``` {r include = FALSE, message = FALSE, echo = FALSE}
library(dplyr)
library(plotly)
library(tidyverse)
library(readxl)
library(lubridate)

plot <- read_xls("buyback.xls")

plot$observation_date <- as.Date(plot$observation_date)

accumulate_by <- function(dat, var) {
  var <- lazyeval::f_eval(var, dat)
  lvls <- plotly:::getLevels(var)
  dats <- lapply(seq_along(lvls), function(x) {
    cbind(dat[var %in% lvls[seq(1, x)], ], frame = lvls[[x]])
  })
  dplyr::bind_rows(dats)
}

f <- plot %>%
  accumulate_by(~observation_date)


d <- ggplot(f, aes(observation_date, frame = frame)) +
  geom_line(aes(y = net_buyback_gdp, colour = "net buybacks"), size = 1) + 
  geom_line(aes(y = corporate_tax_gdp, colour = "corporate tax income"), size = 1) +
  geom_vline(xintercept = as.numeric(plot$observation_date[20]), linetype = "dashed", ) +
  annotate("text", label = "Tax Cut Approved", x = plot$observation_date[15], y = 0, color = "black") +
  labs(y = "Percent of GDP", x = "Date") +
  theme(legend.title = element_blank())
  
```

``` {r include = FALSE, message = FALSE, echo = FALSE}
p <- ggplotly(d, width = 800) %>%
  layout(legend=list(y=0.8, yanchor="top")) %>% 
  add_annotations( text="Colour", xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.8, yanchor="bottom",    # Same y as legend below
                  legendtitle=TRUE, showarrow=FALSE ) %>%
   animation_opts(
    frame = 125,
    easing = "linear",
    transition = 0, 
    redraw = FALSE
  ) %>% 
  animation_slider(
    currentvalue = list(
      prefix = "Date"
    )
  )


```

```{r echo = FALSE}
p
```

Buybacks haven’t just been rising due to the recent tax cut, they have been on the rise since they became legal in ( year of legality) and business investment has been sluggish since it peaked in the early 90s.

<div style='text-align:center'>
<iframe src="https://data.worldbank.org/share/widget?end=2016&indicators=NE.GDI.FTOT.ZS&locations=US&start=2000"&amp;embedded=true" width='650' height='400' align="center" frameBorder='0' scrolling="no" ></iframe>
</div>

Corporate tax cuts are supposed to lead to higher business [investment](https://www.bloomberg.com/view/articles/2017-11-17/yes-a-corporate-tax-cut-would-increase-investment) but instead there’s claims that companies are foregoing investment and using the extra cash to buyback their [shares](http://theweek.com/articles/787966/ban-share-buybacks?utm_source=dlvr.it&utm_medium=twitter). Corporations having more cash at hand and not increasing their rate of investment doesn’t bode well for the economy.This is understandable though since capital expenditure needs careful planning before it’s undertaken and so we may see this rise in the future. Capital expenditure plans are also affected by uncertainty and things like a possible trade war may be delaying plans[Provide links].  



