---
title: "Investment"
author: "Hamdi Ahmed"
date: '2018-09-26'
slug: investment
categories: []
tags: []
---

<style>
/* resize the widget container */
.plotly { 
  width: 100% !important;
}

/* center the widget */
div.svg-container {
  margin: auto !important;
}
</style>


Imagine we find a habitable planet next to our nearest star which wasn’t the sun. That would be Proxima Centauri and it is approximately 4.2 light years away from us. What follows next is a gross oversimplification of space travel. If we assume humans could travel through space at a speed of 56,000 km/h, it would take 81,000 years to reach the planet. We of course decide to colonise it but we can’t live for 81,000 years — the current world average life expectancy is 70.5 years. So, let’s assume our next best idea is to go there by using a generational ship, where this ship would sustain humans for generations until the destination is reached. Those aboard the ship will live and die on it, giving birth to the next generation along the way until the final generation reaches the planet. 

Would we do it? 

81,000 years is obviously a long time and the first humans going aboard the generational ship are making choices based on their current technology. What if 43,587 years later, those back on earth discover new technology which would allow them to colonise the planet faster than the generational ship could, say within 15 years. The offspring of those humans who boarded the generational ship would not reach the planet first, which was the main reason to go on the ship. With this in mind, would anyone choose to board the ship in the first place?

This is the same problem with investment. It takes time for the benefits to show and in that time, it can be rendered useless by new technology. Could this sort of thinking be discouraging businesses from investments?

### **Why is investment important?**

>Investment is central to the functioning of any economy. The process of committing time, resources, and money so that we can produce useful things in the future is, from an economic point of view, a defining part of what businesses, governments, and individuals do.

That’s from Jonathan Haskel’s and Stian Westlake’s book *Capitalism without Capital*. Investment is what allows me to write this post on a laptop instead on pen and paper. It’s what allows me to research this topic on the internet instead of drudging away in the library, sifting through a mountain of books. It’s also what allows you to read this post in your home, place of work or on public transportation without carrying around pieces of paper. 

Investment is a component of GDP — a measure of an economy’s output. It’s the most volatile component as the animation below shows, it rises and falls based on how optimistic firms are about the future. 

``` {r include = FALSE, message = FALSE, echo = FALSE}
  library(readxl)
  library(gifski)
  library(data.table)
  library(lubridate)
  library(stringr)
  library(ggplot2)
  library(dplyr)
  library(tidyverse)
  library(data.table)

mydir <- "C:/Users/Hamdi/Desktop/Website/Hahm/static/images"

volatile2 <-read_excel("volatility2.xlsx")

colnames(volatile2)[1]<-"cycle"

volatile2$year<-substr(volatile2$cycle,1,4)

volatile2$month<-as.numeric(substr(volatile2$cycle,7,7))*3

volatile2$date<-as.Date(ISOdate(volatile2$year,volatile2$month,1))

volatile2 %>% select(-cycle,-year,-month) %>% 
  gather(indicator,percent,-date) ->volatile.df

mycaption<-"Mack, A., and E. Martinez-Garcia. 2011. 'A Cross-Country Quarterly Database of Real House Prices: A Methodological Note.' Globalization and Monetary Policy Institute Working Paper No. 99, Federal Reserve Bank of Dallas."

mycaption<-str_wrap(mycaption,width=80) #wrap the caption

mycap1<-"@lenkiefer Source: Dallas Federal Reserve International House Price Database  http://www.dallasfed.org/institute/houseprice/"  #caption part 2


blist<-c("investment","consumption","gov_expen")

volatile.dt2<-data.table(volatile.df)[year(date)>1970 & indicator %in% blist,]

xlist<-unique(volatile.dt2[year(date)>1979]$date)

M<-length(xlist)


volatile.plot<-function(i=1){
  ggplot(data=volatile.dt2[year(date)>1979 & 
                    date<=xlist[i] &
                    indicator %in% blist],
         aes(x=date,y=percent,color=indicator,linetype=indicator,label=indicator))+
    scale_color_manual(values=c("#CC79A7", "#0072B2", "#D55E00"))+
    geom_line(size=1.1)+
    scale_x_date(breaks=seq(xlist[1],xlist[M]+years(1),"5 year"),
                 date_labels="%Y",limits=c(xlist[1],xlist[M]+years(1)))+
    geom_text(data=volatile.dt2[date==xlist[i] &
                         indicator %in% blist],
              hjust=0,nudge_x=30,size=2.5)+ 
    geom_hline(yintercept=0,linetype=2)+
    scale_y_continuous(breaks=seq(-30,45,5),limits=c(-35,40))+
    theme(legend.position="none",
          plot.caption=element_text(hjust=0,size=7),
          plot.subtitle=element_text(face="italic"))+
    labs(x="",y="",title="Global house price trends through 2018Q1",
         caption=paste0(mycap1,"\n",mycaption),
         subtitle="Seasonally adjusted nominal index (2000Q1=100, log scale)") 
}

```


```{r, animation.hook='gifski'}

volatile.plot(M) 

```

```{r}

files = list.files(
  'content/', '[.](Rmd)$', full.names = TRUE,
  recursive = TRUE
)

```


```{r}

for( f in files ){
  x <- readLines(f)
  y <- gsub( '\\{\\{ site.url\\}\\}', "../../../..", x )
  cat(y, file=f, sep="\n")}


```




```{r}


gif_file <- save_gif({for (i in seq(1,M)){  
  g<- volatile.plot(i)
  print(g)
  print(paste(i,"out of",M))
}
  for (ii in 1:20){
    print(g)
    print(paste(ii,"out of",20))
  }
}, gif_file = paste0(mydir,"/volatile.gif"), width = 1200, height = 800, res = 144, delay=1/10)
utils::browseURL(gif_file)

```



