---
title: "Buybacks & Investment"
author: "Hamdi Ahmed"
date: '2018-08-12'
slug: buybacks-investment
categories: []
tags: []
---

Since the Trump Tax cuts were approved, there's been an expected fall in tax revenues from corporations but there's also been a surge in buybacks.




``` {r include = FALSE, message = FALSE, echo = FALSE}
library(dplyr)
library(plotly)
library(tidyverse)
library(readxl)
library(lubridate)

plot <- read_xls("buyback.xls")

plot$observation_date <- as.Date(plot$observation_date)

accumulate_by <- function(dat, var) {
  var <- lazyeval::f_eval(var, dat)
  lvls <- plotly:::getLevels(var)
  dats <- lapply(seq_along(lvls), function(x) {
    cbind(dat[var %in% lvls[seq(1, x)], ], frame = lvls[[x]])
  })
  dplyr::bind_rows(dats)
}

f <- plot %>%
  accumulate_by(~observation_date)


d <- ggplot(f, aes(observation_date, frame = frame)) +
  geom_line(aes(y = net_buyback_gdp, colour = "net buybacks"), size = 1) + 
  geom_line(aes(y = corporate_tax_gdp, colour = "corporate tax income"), size = 1) +
  geom_vline(xintercept = as.numeric(plot$observation_date[20]), linetype = "dashed", ) +
  annotate("text", label = "Tax Cut Approved", x = plot$observation_date[15], y = 0, color = "black") +
  labs(y = "Percent of GDP", x = "Date") +
  theme(legend.title = element_blank())
  
```

``` {r include = FALSE, message = FALSE, echo = FALSE}
p <- ggplotly(d, width = 800) %>%
  layout(legend=list(y=0.8, yanchor="top")) %>% 
  add_annotations( text="Colour", xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.8, yanchor="bottom",    # Same y as legend below
                  legendtitle=TRUE, showarrow=FALSE ) %>%
   animation_opts(
    frame = 100,
    easing = "linear",
    transition = 0, 
    redraw = FALSE
  ) %>% 
  animation_slider(
    currentvalue = list(
      prefix = "Date"
    )
  )


```

```{r echo = FALSE}
p
```

It's fine though since corporate tax [cuts](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1091238) are supposed to [lead](https://twitter.com/realdonaldtrump/status/942123433873281024?lang=en) to [higher](https://www.bloomberg.com/view/articles/2017-11-17/yes-a-corporate-tax-cut-would-increase-investment) business investment, but there’s claims that companies are foregoing investment and using the extra cash to buyback their [shares](http://theweek.com/articles/787966/ban-share-buybacks?utm_source=dlvr.it&utm_medium=twitter). Indeed, buybacks haven’t just been rising due to the recent tax cut, they have been on the rise since they became legal in 1982 and corporate capital investment, although cyclical, has been sluggish.


``` {r include = FALSE, message = FALSE, echo = FALSE}

library(zoo)
library(tidyr)


invest <- read_xlsx("invest.xlsx")

invest$year <- as.Date(as.yearqtr(invest$year))



m <- ggplot(invest, aes(x = year)) +
  geom_line(aes(y = net_buybacks, colour = "Net buybacks")) + 
  geom_line(aes(y = investment, colour = "U.S Investment")) +
  labs(y = "Percent of GDP", x = "Year") 

```

```{r echo = FALSE, fig.align='center'}

m

```
**( Talk about the numbers in this graph, compare average investment/ buyback in later decades to average investment/buybacks in recent decades. Then talk more about general fall in the pace of investment in recent years, maybe talk about what type of industries this fall in recent investment is concentrated in)**

U.S net buyback as a percentage of GDP from 1960-1985 averaged -0.09. It makes sense for corporations to issue more shares than they bought back in that period as the act of buying back shares was only legal for 3 years. From 1985-2005 this number did average 0.97, that's a percentage increase of approximately 1176%. For investment as a share of GDP, its percentage increase was just approximately 1.56%, averaging 8.86 in the period 1960-1985 and 9 in the period 1985-2005. So, it looks like corporations are increasing their rate of buybacks faster than their investments.

If corporations having more cash at hand doesn't lead to an increase in their investments, that won't bode well for the economy. This is understandable though since capital expenditure needs to be planned before it’s undertaken and so we may see this rise in the future. Capital expenditure plans are also affected by uncertainty and things like a possible trade war doesn't [help](https://www.telegraph.co.uk/business/2018/06/20/us-businesses-putting-investment-trade-war-worries-fed-chief/). That being said though, if a company is buying back more shares then there’s less resources available to devote to investment right?  John Cochrane says otherwise as companies could be giving back cash to shareholders because of a lack of profitable projects to [invest in](https://johnhcochrane.blogspot.com/2017/12/the-buyback-fallacy.html).

>If company 1 doesn't have any good investment ideas, even after the tax cut, and company 2 does have some good investment ideas, made better after the tax cut, the economy needs to get money from company 1 to company 2. Company 1 could buy company 2; company 1 could invest in company 2 by buying its stock or buying its debt (all that "cash" you hear about); company 1 could return money to shareholders, and the shareholders could invest in company 2. They're all the same, to economics. Of all the ways to do this, actually, the last might well be the most efficient. Shareholders might have better ideas about good investments than managers of a company that doesn't have any good investment ideas.

It becomes an issue though if most companies are handing back cash to [shareholders](https://www.bloomberg.com/view/articles/2018-07-18/trump-s-tax-cut-hasn-t-done-anything-for-workers). Is this a sign of a lack of good investment projects available in the economy or would companies be investing more if they weren’t distracted by the ability to buyback shares?

First, some definitions. Share buybacks or share repurchases is the act of returning cash to shareholders by buying stock held by them. With less shares available, the remaining investors get a bigger piece of corporate profits. However, corporations may issue new shares in the same time period as they buy back shares, so I use net share buybacks in my graphs to take this into account. Net share buybacks also show us the amount that is left over for investment. I use investment defined as gross fixed capital formation undertaken by the non-financial corporate sector. Gross [fixed capital](https://en.wikipedia.org/wiki/Fixed_capital) formation measures the net increase in fixed capital investment which is a component of the [expenditure method in calculating GDP](http://www.econport.org/content/handbook/NatIncAccount/CalculatingGDP/Expenditures.html). Both data sets are from the [Federal Reserve](https://www.federalreserve.gov/econres/notes/ifdp-notes/corporate-buybacks-and-capital-investment-an-international-perspective-20170411.htm).

### **Do share buybacks deter investment?**

The point made by those that say yes is that management are deliberately forgoing investment because they incentivised to buyback shares instead. 

The point made by those that say yes is that corporate executives are incentivised to buyback shares above all else. This is [because](https://hbr.org/2014/09/profits-without-prosperity):

>Stock-based instruments make up the majority of their pay, and in the short term buybacks drive up stock prices.

This Deloitte [article](https://www2.deloitte.com/insights/us/en/economy/behind-the-numbers/corporate-share-buybacks-business-investment.html#endnote-sup-3) explains how:

>A share buyback reduces the number of shares in the market and therefore increases the price and earnings per share of the remaining outstanding shares.


There is evidence that share [buybacks]( https://papers.ssrn.com/sol3/papers.cfm?abstract_id=250049) do [raise]( https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2664098) share prices. But this evidence has less to do with buybacks cutting the number of shares available and instead supports this [explanation](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3082460) by Cliff Asness et al:

>First, repurchases might signal that management believes that shares are undervalued…Second, because interest payments are tax deductible, debt financed repurchases can be viewed as good news due to the resulting lower tax burden. Third, investors may feel as though it is better for management to return excess cash to shareholders, rather than chasing less economic “pet” projects. This kind of agency cost is often characterized as “empire building,” and avoiding it has long been viewed as one of the benefits of returning cash to shareholders.


### **TLDR**

All of this shows that the negative effect of buybacks on investment is weak at best and there are stronger forces at work. What could these be? I expore them in my next post. Thank you for reading.
